Assessing Bridge Condition in New York State Using Multiple Linear Regression

### 2024-12-04 paper/paper.qmd
Improve prose

---
title: "Assessing Bridge Condition in New York State Using Multiple Linear Regression"
subtitle: "Age, Municipality, and Ownership as Factors Influencing Bridge Condition Ratings "
author: 
  - Ariel Xing
thanks: "Code and data are available at: [https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis](https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis)."
date: today
date-format: long
abstract: "This study examines the factors influencing bridge conditions in New York State, focusing on the roles of age, municipality location, and ownership, using condition ratings provided by NYSDOT as a measure of bridge quality. Applying a multiple linear regression model, the analysis reveals that for every additional year of age, a bridge's condition rating decreases by 0.02 points, reflecting the progressive impact of aging. On average, bridges located in towns and villages have condition ratings that are 0.17 and 0.19 points higher, respectively, than those in cities. Additionally, bridges owned by NYSDOT and other entities have condition ratings that are 0.09 and 0.16 points lower, respectively, than those owned by municipalities. These findings highlight the importance of targeted maintenance strategies that account for geographic and administrative factors to enhance infrastructure sustainability and safety.
"
format: pdf
toc: true 
number-sections: true
bibliography: references.bib
---

@@

This dataset was chosen over the updated 2020 version from New York State Open Data due to the change in how bridge condition data is recorded. In the 2016 dataset, the condition variable is a continuous variable measured on a seven-point scale (1 = "Deficient Condition" to 7 = "Good Condition"). However, in the 2020 dataset, NYSDOT replaced this scale with a binary Poor/OK variable, recorded as Poor01 (1 = Poor, 0 = OK). This change limits the analytical potential of the dataset, as the binary variable lacks the granularity required for linear regression that relies on continuous variables to detect nuanced trends or incremental differences. By using the 2016 dataset, this study retains the detailed numerical condition ratings, enabling to analysis impact of factors such as age, location, and owners on bridge conditions.

@@

## Outcome Variable
### Condition 
The **Condition** variable reflects the overall state of a bridge, encompassing both its structural integrity and functional adequacy as determined during inspections. It is measured on a continuous scale from 1 to 7, with higher scores indicating better condition. Bridges with condition scores above 5 are considered to be in good standing, while scores below 5 indicate that a bridge is classified as either Structurally Deficient (SD) or Functionally Obsolete (FO).

@@

## Model Justification
### Variable Choice
The selection of predictors in the model reflects key aspects discussed in the data section, ensuring a comprehensive understanding of bridge condition ratings. Age at Inspection was included as a continuous variable because @fig-ageatinspection-condition demonstrated a clear relationship between bridge age and condition ratings. Using age as a continuous predictor avoids the loss of information that would occur if age were categorized into groups, allowing the model to capture subtle variations in condition ratings as bridges age. Additionally, this choice aligns with the dataset's structure, where age is measured on a continuous scale.

Municipality Type (City, Town, Village) was treated as a categorical variable with Town as the reference category. The data revealed that condition ratings vary systematically between municipality types, likely due to differences in traffic loads, maintenance resources, and environmental stressors. Modeling Municipality Type as levels rather than using dummy variables for individual municipalities ensures clarity and interpretability, highlighting differences between City, Town, and Village bridges.

@@

### Limitations
Several limitations must be acknowledged. Due to data constraints, the model excludes potentially influential factors such as transport load, weather conditions, and environmental influences, introducing the risk of omitted variable bias. This limitation reduces the explanatory power of the model and may skew the estimated effects of included predictors. Additionally, the model does not account for interaction effects, such as how age and municipality might jointly influence condition ratings. Incorporating such interactions could enhance the model’s explanatory ability but would increase its complexity. Finally, the linearity assumption, while reasonable, may oversimplify more nuanced relationships in the data.

@@

## Age Distribution Patterns {#sec-age-distribution}
The distribution of bridge ages at inspection reveals distinct peaks approximately every 25 years, indicating a recurring cycle of bridge construction and major renovation. This pattern suggests that every 25 years, a significant number of bridges reach the point where extensive maintenance or replacement is necessary. The previous peak occurred roughly 25 years ago, meaning the next wave of large-scale repairs or new construction is approaching. Recognizing this pattern highlights the importance of early planning and budgeting to address the anticipated surge in infrastructure needs effectively, ensuring resources are ready to manage the simultaneous demands of aging bridges.

## Repair Prioritization and Budget Allocation
The regression model highlights the critical factors of bridge age, municipality type, and ownership in determining condition ratings. Urban bridges and those owned by NYSDOT or private entities consistently exhibit lower condition ratings, likely due to heavier traffic loads and diverse maintenance practices. These findings underscore the need for targeted repair strategies. Municipalities, which generally achieve better condition ratings, could serve as a benchmark for best practices in resource utilization. Repair prioritization should integrate both structural condition and traffic intensity to maximize the effectiveness of budget allocations, ensuring that high-use and socioeconomically vital bridges receive timely attention.

@@

## Geographic and Administrative Disparities in Bridge Conditions
The analysis reveals significant disparities in bridge conditions based on geographic location and administrative ownership. Bridges in urban areas tend to have lower condition ratings compared to those in towns and villages, likely due to higher traffic volumes, environmental stress, and more complex structural demands in urban settings. Similarly, bridges owned by NYSDOT or private entities perform worse than those managed by municipalities, suggesting differences in maintenance priorities and resource allocation. These disparities point to the need for a more equitable distribution of infrastructure funds and tailored maintenance strategies that address the specific challenges faced by different regions and ownership groups. For example, urban bridges may benefit from more frequent inspections and targeted upgrades to withstand heavy usage, while state-owned and privately managed bridges could require enhanced funding mechanisms or stricter maintenance policies to ensure their long-term sustainability. Recognizing and addressing these geographic and administrative variations will be critical for maintaining a resilient and equitable bridge infrastructure network.

@@

## Overview
The objective of this methodology is to design a comprehensive system for collecting and analyzing traffic flow data across New York State bridges. This additional data addresses a critical gap in the current dataset, which does not include traffic intensity—a key factor influencing bridge deterioration. By integrating traffic flow data, the enhanced model aims to improve its explanatory power and provide actionable recommendations for bridge maintenance prioritization. Rigorous sampling strategies and advanced data collection techniques are employed to ensure that the data is both representative and reliable, supporting more nuanced policy and engineering decisions.

## Target Population
The target population encompasses all operational bridges in New York State. These include bridges in urban, suburban, and rural areas, varying by traffic volume and ownership. Bridges were selected to account for:

@@

By incorporating these dimensions, the target population captures the variability in bridge conditions and usage, enabling a comprehensive analysis of how traffic impacts structural integrity.

@@

### Manual Surveys
- Description: Periodic on-site observations validate automated systems and address gaps in data, especially for rural or low-traffic bridges.
- Advantages: Provides ground-truth data for calibration.
- Limitations: Resource-intensive and limited to periodic snapshots.
## Data Validation and Quality Control.

@@

### Budget Constraints
- Issue: Advanced systems, such as AI-enabled cameras, require significant investment, potentially limiting their use.
- Mitigation: Employ a hybrid approach combining cost-effective methods (AVCs, GPS) with targeted deployment of high-cost systems.

>>>>

---
title: "Assessing Bridge Condition in New York State Using Multiple Linear Regression"
subtitle: "Age, Municipality, and Ownership as Factors Influencing Bridge Condition Ratings "
author: 
  - Ariel Xing
thanks: "Code and data are available at: [https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis](https://github.com/Jiaqi-Xing/NYS_Bridge_Condition_Analysis)."
date: today
date-format: long
abstract: "This study examines the factors influencing bridge conditions in New York State, focusing on the roles of age, municipality location, and ownership, using condition ratings provided by NYSDOT as a measure of bridge quality. Applying a multiple linear regression model, the analysis shows that for every additional year of age, a bridge's condition rating decreases by 0.02 points, reflecting the progressive impact of aging. On average, bridges located in towns and villages have condition ratings that are 0.17 and 0.19 points higher, respectively, than those in cities. Additionally, bridges owned by NYSDOT and other entities have condition ratings that are 0.09 and 0.16 points lower, respectively, than those owned by municipalities. These findings highlight the importance of targeted maintenance strategies that account for geographic and administrative factors to enhance infrastructure sustainability and safety.
"
format: pdf
toc: true 
number-sections: true
bibliography: references.bib
---

@@

This dataset was chosen over the updated 2020 version from New York State Open Data due to the change in how bridge condition data is recorded. In the 2016 dataset, the condition variable is a continuous variable measured on a seven-point scale (1 = "Deficient Condition" to 7 = "Good Condition"). However, in the 2020 dataset, NYSDOT replaced this scale with a binary Poor/OK variable, recorded as Poor01 (1 = Poor, 0 = OK). This change limits the analytical potential of the dataset, as the binary variable lacks the granularity required for linear regression that relies on continuous variables to detect subtle trends or incremental differences. By using the 2016 dataset, this study retains the detailed numerical condition ratings, enabling to analysis impact of factors such as age, location, and owners on bridge conditions.

@@

## Outcome Variable
### Condition 
The **Condition** variable reflects the overall state of a bridge, capturing both its structural integrity and functional adequacy as determined during inspections. It is measured on a continuous scale from 1 to 7, with higher scores indicating better condition. Bridges with condition scores above 5 are considered to be in good standing, while scores below 5 indicate that a bridge is classified as either Structurally Deficient (SD) or Functionally Obsolete (FO).

@@

## Model Justification
### Variable Choice
The selection of predictors in the model reflects key aspects discussed in the data section, ensuring a thorough understanding of bridge condition ratings. Age at Inspection was included as a continuous variable because @fig-ageatinspection-condition demonstrated a clear relationship between bridge age and condition ratings. Using age as a continuous predictor avoids the loss of information that would occur if age were categorized into groups, allowing the model to capture subtle variations in condition ratings as bridges age. Additionally, this choice aligns with the dataset's structure, where age is measured on a continuous scale.

Municipality Type (City, Town, Village) was treated as a categorical variable with Town as the reference category. The data shows that condition ratings vary systematically between municipality types, likely due to differences in traffic loads, maintenance resources, and environmental stressors. Modeling Municipality Type as levels rather than using dummy variables for individual municipalities ensures clarity and interpretability, highlighting differences between City, Town, and Village bridges.

@@

### Limitations
Several limitations must be acknowledged. Due to data constraints, the model excludes potentially influential factors such as transport load, weather conditions, and environmental influences, introducing the risk of omitted variable bias. This limitation reduces the explanatory power of the model and may skew the estimated effects of included predictors. Additionally, the model does not account for interaction effects, such as how age and municipality might jointly influence condition ratings. Incorporating such interactions could enhance the model’s explanatory ability but would increase its complexity. Finally, the linearity assumption, while reasonable, may oversimplify more subtle relationships in the data.

@@

## Age Distribution Patterns {#sec-age-distribution}
The distribution of bridge ages at inspection shows distinct peaks approximately every 25 years, indicating a recurring cycle of bridge construction and major renovation. This pattern suggests that every 25 years, a significant number of bridges reach the point where extensive maintenance or replacement is necessary. The previous peak occurred roughly 25 years ago, meaning the next wave of large-scale repairs or new construction is approaching. Recognizing this pattern highlights the importance of early planning and budgeting to address the anticipated surge in infrastructure needs effectively, ensuring resources are ready to manage the simultaneous demands of aging bridges.

## Repair Prioritization and Budget Allocation
The regression model highlights the key factors of bridge age, municipality type, and ownership in determining condition ratings. Urban bridges and those owned by NYSDOT or private entities consistently exhibit lower condition ratings, likely due to heavier traffic loads and diverse maintenance practices. These findings underscore the need for targeted repair strategies. Municipalities, which generally achieve better condition ratings, could serve as a benchmark for best practices in resource utilization. Repair prioritization should integrate both structural condition and traffic intensity to maximize the effectiveness of budget allocations, ensuring that high-use and socioeconomically vital bridges receive timely attention.

@@

## Geographic and Administrative Disparities in Bridge Conditions
The analysis shows significant disparities in bridge conditions based on geographic location and administrative ownership. Bridges in urban areas tend to have lower condition ratings compared to those in towns and villages, likely due to higher traffic volumes, environmental stress, and more complex structural demands in urban settings. Similarly, bridges owned by NYSDOT or private entities perform worse than those managed by municipalities, suggesting differences in maintenance priorities and resource allocation. These disparities point to the need for a more equitable distribution of infrastructure funds and tailored maintenance strategies that address the specific challenges faced by different regions and ownership groups. For example, urban bridges may benefit from more frequent inspections and targeted upgrades to withstand heavy usage, while state-owned and privately managed bridges could require enhanced funding mechanisms or stricter maintenance policies to ensure their long-term sustainability. Recognizing and addressing these geographic and administrative variations will be essential for maintaining a resilient and equitable bridge infrastructure network.

@@

## Overview
The objective of this methodology is to design a thorough system for collecting and analyzing traffic flow data across New York State bridges. This additional data addresses an essential limitation in the current dataset, which does not include traffic intensity—a key factor influencing bridge deterioration. By integrating traffic flow data, the enhanced model aims to improve its explanatory power and provide actionable recommendations for bridge maintenance prioritization. Rigorous sampling strategies and sophisticated data collection techniques are employed to ensure that the data is both representative and reliable, supporting more detailed policy and engineering decisions.

## Target Population
The target population includes all operational bridges in New York State. These include bridges in urban, suburban, and rural areas, varying by traffic volume and ownership. Bridges were selected to account for:

@@

By incorporating these dimensions, the target population captures the variability in bridge conditions and usage, enabling a detailed analysis of how traffic impacts structural integrity.

@@

### Manual Surveys
- Description: Periodic on-site observations validate automated systems and address missing information in data, especially for rural or low-traffic bridges.
- Advantages: Provides ground-truth data for calibration.
- Limitations: Resource-intensive and limited to periodic snapshots.
## Data Validation and Quality Control.

@@

### Budget Constraints
- Issue:  High-cost systems, such as AI-enabled cameras, require significant investment, potentially limiting their use.
- Mitigation: Employ a hybrid approach combining cost-effective methods (AVCs, GPS) with targeted deployment of high-cost systems.
